Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2012-01-06 article.tex
Brought discussion of Q factor back to sampling theorem; mentioned weighted SVD
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1451 ccdd05d5-56c1-461a-bf98-85f6876523b4

One possible source of \SNR\ loss for which we have not accounted is the leakage of sharp spectral features built into the whitening filter.  In the \lloid\ algorithm, as with many other \GW\ search methods, the whitening procedure is treated as a data conditioning procedure that is separated from the matched filter stage.  In this paper, we assume that the input to the filter bank is a whitened data stream --- the result of feeding the \GW\ antenna's strain time series through a filter that flattens and normalizes its spectrum --- and how the whitening filter is implemented is not described in detail.  We elected to omit a detailed description of the whitening procedure since the focus here is on the implementation of a scalable inspiral filter bank.

@@

In order to effectively flatten a line in the detector's noise power
spectrum, the time scale of the templates must be at least as long as
the damping time $\tau$ of the line, $\tau = 2 Q / \omega_0$, where $Q$ is the quality
factor of the line and $w_0$ is the central angular frequency.  For example,
in the S6 detector configuration known as ``Enhanced LIGO,'' the violin modes \citep{ELIGOSusp} had $Q \sim 10^5$ and $\omega_0 \sim (2 \pi) 340$~rad/s, for a coherence time $\tau \sim 10^2$~s.

In our example template bank, many of the time slices are much shorter
than this.  However, the impulse response of the LLOID filter bank is very nearly
equal to the full templates themselves, and the full templates are much longer than many
coherence times of the violin mode.  For this reason, we do not expect
\lloid\ to be affected by sharp line features any more than the
traditional \fft-based searches currently employed in the \GW\ field.  Future works ought to verify this reasonable expectation with numerical experiments.

There could, in principle, be lines with coherence times many times
longer than the template duration.  For example, the $Q$ of the violin
modes may increase by orders of magnitude in Advanced \LIGO~\citep{ALIGOSusp}.  Also,
there are certainly narrow lines that are non-stationary.  Both of these
cases can be dealt with by preprocessing $h(t)$ with bandstop filters
that attenuates the lines themselves but also conservative neighborhoods
around them.  If such a bandstop was implemented as an \fir\ filter, it
could be built into the time slices just as before.

Another way to deal with line features with coherence times much longer than the templates would be to entrirely `factor' the whitening out of the \lloid\ filter bank.  The data would be passed through the whitening filter twice and the inspiral templates would not depend on the detector's noise power spectral density at all.  Any line features could be notched out in the whitening stage with IIR filters, which can achieve infinitely high $Q$ at just second order.

>>>>

One possible source of \SNR\ loss for which we have not accounted is the leakage of sharp spectral features in the detector's noise spectrum due to the short durations of the time slices.  In the \lloid\ algorithm, as with many other \GW\ search methods, whitening is treated as an entirely separate data conditioning stage.  In this paper, we assume that the input to the filter bank is already whitened, having been passed through a filter that flattens and normalizes its spectrum.  We elected to omit a detailed description of the whitening procedure since the focus here is on the implementation of a scalable inspiral filter bank.

@@

In order to effectively flatten a line in the detector's noise power
spectrum, the time scale of the templates must be at least as long as
the damping time $\tau$ of the line, $\tau = 2 Q / \omega_0$, where $Q$ is the quality
factor of the line and $w_0$ is the central angular frequency.  To put this into the context of the sampling theorem, in order to resolve a notch with a particular $Q$ and $f_0$, a \fir\ filter must achieve a frequency resolution of $\Delta f \gtrsim \pi f_0 / Q$ and therefore its impulse response must last for at least a time $1/\Delta f = Q / \pi f_0$.  For example,
in the S6 detector configuration known as ``Enhanced LIGO,'' the violin modes \citep{ELIGOSusp} had $Q \sim 10^5$ and $\omega_0 \sim (2 \pi) 340$~rad/s, for a coherence time $\tau \sim 10^2$~s.

In our example template bank, many of the time slices are much shorter
than this.  However, in summation the time slices have the same duration as the full templates themselves, and the full templates are much longer than many
coherence times of the violin mode.  For this reason, we speculate that 
\lloid\ should be just as robust to sharp line features as
traditional \fft-based searches currently employed in the \GW\ field.  Future works must verify this reasonable supposition with numerical experiments, including impulse response studies similar to the ones presented here but with detector noise power spectra containing lines with realistically high quality factors.

There could, in principle, be lines with coherence times many times
longer than the template duration.  For example, the $Q$ of the violin
modes may increase by orders of magnitude in Advanced \LIGO~\citep{ALIGOSusp}.  Also,
there are certainly narrow lines that are non-stationary.  Both of these
cases can be dealt with by preprocessing $h(t)$ with bandstop filters
that attenuates the lines themselves but also conservatively large neighborhoods
around them.  If such bandstops were implemented as an \fir\ filter, they
could be built into the time slices without any difficulty.

Perhaps another way to deal with line features with coherence times much longer than the templates would be to entrirely `factor' the whitening out of the \lloid\ filter bank.  Any line features could be notched out in the whitening stage with IIR filters, which can achieve infinitely high $Q$ at just second order.  If the detector data were passed through the whitening filter twice, then time-sliced filters need not depend on the detector's noise power spectral density at all.  In such a variation on the \lloid\ method, the basis filters could be calculated from the \emph{weighted} \SVD~\citep[chapter 3.6]{WeightedSVD, jackson2003user} of the time-sliced templates, using the covariance of the detector noise as a weight matrix.
